---
title: "Data607_Project2 - Electricity"
author: "Ryan Weber"
date: "March 11, 2018"
output:
  html_document:
      theme: cerulean
      highlight: pygments
---

```{r setup, include=FALSE}
rm(list=ls())
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

Files and data stored at: https://github.com/rweberc/Data607_Project2

# Electricity demand (contributed by Rose Koh)

Net actual interchange (electric):  The algebraic sum of all metered interchange over all interconnections between two physically Adjacent Balancing Authority Areas. NERC definition
https://www.eia.gov/tools/glossary/?id=electricity

### Read in data

```{r}
elecDf <- read.csv("https://raw.githubusercontent.com/rweberc/Data607_Project2/master/electricity.csv", stringsAsFactors = FALSE, colClasses = "character", skip = 4)
```

### Confirm data structure:
- Every second line "Demand"
- Every third line ""
- Every fourth line "Total net actual interchange"
```{r}
stopifnot(all(elecDf$megawatthours[c(FALSE, TRUE, FALSE, FALSE)] == "Demand"))
stopifnot(all(elecDf$megawatthours[c(FALSE, FALSE, TRUE, FALSE)] == "Net generation"))
stopifnot(all(elecDf$megawatthours[c(FALSE, FALSE, FALSE, TRUE)] == "Total net actual interchange"))
```

### Extract region names
```{r}
elecDf$region <- NA
elecDf$region[c(TRUE, FALSE, FALSE, FALSE)] <- elecDf$megawatthours[c(TRUE, FALSE, FALSE, FALSE)]

elecDf <- elecDf %>% fill(region)
table(elecDf$region, useNA = "ifany")

elecDf$region[c(TRUE, FALSE, FALSE, FALSE)] <- ""
```

### Update column names
```{r}
elecDf <- elecDf %>% rename(category = megawatthours)
elecDf <- elecDf %>% select(category, region, everything())

colnames(elecDf) <- str_replace(colnames(elecDf), "X", "")
colnames(elecDf) <- str_replace_all(colnames(elecDf), "\\.", "/")

colnames(elecDf)
```

### Wide to long
```{r}
elecDf_long <- elecDf %>% gather("evalDay", "megawatthours", 3:length(.))
```

### Clean data in long form
```{r}
# data check
elecDf_long$megawatthours[elecDf_long$megawatthours == " Inc. (region)"] <- ""
stopifnot(elecDf_long$megawatthours[str_trim(elecDf_long$region) == ""] == "")

# filter out unneed "region" rows
elecDf_long <- elecDf_long %>% filter(str_trim(region) != "")

# parse day and megawatthours
elecDf_long$evalDay <- mdy(elecDf_long$evalDay)
elecDf_long$megawatthours <- as.numeric(elecDf_long$megawatthours)
```

### Long to wide
```{r}
elecDf <- elecDf_long %>% spread(category, megawatthours)
rm(elecDf_long)

# data check: confirm no megawatt hours are NA
stopifnot(nrow(elecDf %>% filter_if(is.Date, any_vars(is.na(.)))) == 0)
```

### Suggested analyses
*  daily demand per region
*  daily net generation per region
*  daily total net actual interchange
```{r}
perRegionAvg <- elecDf %>% 
  group_by(region) %>%
  summarize(demandAvg = mean(Demand),
                                     netGenAvg = mean(`Net generation`),
                                     totInterchgAvg = mean(`Total net actual interchange`))

ggplot(perRegionAvg, aes(x=region, y=totInterchgAvg)) + 
  geom_bar(stat = "identity") +
  labs(title = "Avg Interchange values", x = "Region", y = "Averge Interchange") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0))

```

California looks to have the greatest deficits, while the northwest region looks to have the greatest surplus.

*  Any particular days were demand was greater than production?
```{r}
dayDf <- elecDf %>% 
  ungroup() %>%
  mutate(evalDayOfWeek = wday(evalDay, label = TRUE)) %>%
  group_by(region, evalDayOfWeek) %>%
  summarize(demandAvg = mean(Demand),
                                     netGenAvg = mean(`Net generation`),
                                     totInterchgAvg = mean(`Total net actual interchange`))

ggplot(dayDf, aes(x = region, y = totInterchgAvg)) + 
  geom_bar(aes(fill = evalDayOfWeek), position = "dodge", stat = "identity") +
  labs(title = "Avg Interchange values", x = "Region", y = "Averge Interchange") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust = 0))
```

Net values do not seem dependent on days.
